How topos theory can help commutative algebra

### 2018-07-04 slides-como2018.tex
More notes

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}

@@

  The metatheorem soups up a number of lemmas of algebraic geometry, there
  stated in geometric language. For instance, if~$M$ is finitely generated,
  then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.}

@@

  \jnote{3}{XXX}

@@

>>>>

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\renewcommand{\AA}{\mathbb{A}}
\renewcommand{\C}{\mathcal{C}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ull}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ull{\AA}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}{\vdash_{\!\!\!\vec x}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates

@@

  The metatheorem soups up a number of lemmas of algebraic geometry, there
  stated in geometric language. For instance, if~$M$ is finitely generated,
  then~$M^\sim$ is of finite type. If~$M$ is finitely presented, then~$M^\sim$
  is of finite presentation. If~$M$ is coherent, then~$M^\sim$ is coherent.

  As an aside, the little Zariski topos is seldomly Boolean (validates the law of
  excluded middle). A necessary condition is that~$A$ is of dimension~$\leq 0$.}

@@

  \jnote{3}{Here's a rough sketch why the double negations occur. In
  undergraduate linear algebra, we learn that any finitely generated vector
  space is free (has a basis), by the following argument:
  Let~$(v_1,\ldots,v_n)$ be a generating family. Either one of the generators can
  be expressed as a linear combination of the others or not. In the latter
  case, the family is linearly independent and we're done. In the former, we
  remove the redundant generator and continue by induction.

  This argument relies on the law of excluded middle and can therefore not put
  to work in the internal universe as is. However, intuitionistically we do
  have the weaker statement~$\neg\neg(\varphi \vee \neg\varphi)$, and we can
  use this law in place of the law of excluded middle to intuitionistially
  deduce that any finitely generated vector space does \emph{not not} have a
  basis.}

@@

  \jnote{1}{At the
  \fixedhref{https://sbseminar.wordpress.com/2009/08/06/algebraic-geometry-without-prime-ideals/}{Secret
  Blogging Seminar}, there was an insightful long-running discussion on the
  merits of the functorial approach. Two disadvantages of the approach using
  locally ringed spaces is that the underlying topological spaces don't
  actually parametrize ``honest'', ``geometric'' points, but irreducible closed
  subsets; and that they don't work well in a constructive setting. (For this,
  they would have to be replaced by locally ringed locales.)

  Given a functor~$F : \mathrm{Ring} \to \Set$, we imagine~$F(A)$ to be the set
  of~``$A$-valued points'' of the hypothetical scheme described by~$F$, the set
  of ``points with coordinates in~$A$''. These sets are geometrically important.

  The functorial approach is more economical, philosophically rewarding, and
  works constructively. However, typically only field-valued points are easy
  to describe. For instance, the functor representing projective~$n$-space is
  given on fields by
  \[ \begin{array}{r@{}c@{}l}
  K &{}\longmapsto{}& \text{the set of lines through the origin in~$K^{n+1}$} \\
  && \qquad \cong \{ [x_0 \hg \cdots \hg x_n] \,|\, \text{$x_i \neq 0$ for some~$i$} \},
  \end{array} \]
  whereas on general rings it is given by
  \[ \begin{array}{r@{}c@{}l}
    A &{}\longmapsto{}& \text{the set of quotients~$A^{n+1} \twoheadrightarrow P$,
    where~$P$ is projective,} \\
    && \qquad \text{modulo isomorphism}.
  \end{array} \]
  It are these more general kinds of points which impart a meaningful sense of
  cohesion on the field-valued points, so they can't simply be dropped from
  consideration.}

  \jnote{2}{This tension is resolved by observing that the category of
  functors~$\Ring \to \Set$ is a topos (the big Zariski topos of~$\Spec(\ZZ)$)
  and that we can therefore employ its internal language. This language takes
  care of managing stages behind the scenes. For instance, projective~$n$-space
  can be described by the naive expression
  \[ \{ (x_0,\ldots,x_n) : (\affl)^{n+1} \,|\, x_0 \neq 0 \vee \cdots \vee x_n
  \neq 0 \}/(\affl)^\times. \]
  This example illustrates our goal: to develop a synthetic account of
  algebraic geometry, in which schemes are plain sets and morphisms between
  schemes are maps between those sets. It turns out that there are many
  similarities with the well-developed synthetic account of differential
  geometry, but also important differences.}
